Composable Generalized Consensus

### 2015-05-15 paper/paper.tex
Work on presenting the Simple-Leader algorithm.

\subsection{Fast Generalized Consensus}

In our first algorithm, Fast GC, learners each maintain a variable $hist\left[ l \right]$ which contain the currently accepted history of learner $l$.
For every learner $l$, $accepted\left[ l \right]$ is initialized to the empty history.
A proposer broadcasts its proposal to all the learners which unconditionally accept it by setting appending it to their $accepted\left[ l \right]$
variable and broadcasting to all the acceptors the new value of their accepted history.
Each acceptor $l$ maintains a variable $learned\left[ l \right]$, initially the empty history, containing its last learned history. A learner $l$
learns a new history $h$ when $h \geq learned\left[ l \right]$ and $h$ is a lower bound of the set consisting of the last accepted history received
from each acceptor.

@@

\item A $Learn\left( l, h \right)$ transition, for a learner $l$ and a history $h$, enabled when $h \geq learned\left[ l \right]$ and $h$ is a lower
bound of the set $\aset{LastReceived\left( a \right) : a \in Acceptors}$, where $LastReceived\left( a \right)$ is the last accepted history received
from the acceptor $a$.

@@

The algorithm Fast GC cannot make progress if two acceptors $a_1$ and $a_2$ have two histories which are not compatible because, since one can only append to the histories $accepted\left[ l \right]$, all the possible future lower bounds of the set $\aset{LastReceived\left( a \right) : a \in Acceptors}$ are bounded above by $GLB\left( \aset{accepted\left[ a_1 \right], accepted\left[ a_2 \right]} \right)$.
This situation occurs when there is contention on non-commuting commands: for example, when two different proposers send two commands $c_1$ and $c_2$ that do not commute, an acceptors $a_1$ accepts $c_1$ before $c_2$, and an acceptor $a_2 \neq a_1$ accepts $c_2$ before $c_1$.

@@

We write $FGC_1$ for the first, deadlocked, round of the Fast GC algorithm.
The rounds $FCG_1$ and $FCG_2$ each have their own copies of the arrays $accepted$ and $learned$, noted $accepted_1$, $accepted_2$, $learend_1$, and $learned_2$ when the round is not clear from the context. 
We assume that network messages are tagged with a round number and that a message tagged with $i\in\aset{1,2}$ can only be received in $FCG_i$, creating two isolated logical networks.

@@

We modify the FCG algorithm by introducing three new transitions and modifying the $Accept\left( a \right)$ transition as follows.
We add the new variables $status\left[ a \right]$ for every acceptor $a$, which are initialized to $\quo{idle}$. The purpose of $status\left[ a \right]$ is to prevent the acceptor $a$ from accepting commands until its status changes from $\quo{idle}$ to $\quo{ready}$.
We also add the new variables $inits\left[ s \right]$ and $aborts\left[ s \right]$, for every switcher $s$, initialized to the empty history in $FGC_1$ and to $\quo{none}$ in $FGC_2$. The purpose of the arrays $aborts$ and $inits$ is to transfer state between $FGC_1$ and $FGC_2$.
The algorithms $FGC_1$ and $FGC_2$ each have their own copy of the arrays $status$, $inits$, and $aborts$, and we adopt the same convention as above consisting in adding a subscript denoted the round to a variable when the round is not clear from the context.

\begin{compactitem}
  \item We modify the $Accept\left( a \right)$ transition by having the acceptor $a$ additionally broadcast its new accepted value to the switchers.
Switchers can therefore detect that the algorithm is deadlocked when there are two broadcast messages $\aseq{\quo{accepted}, a_1, h_1}$ and $\aseq{\quo{accepted}, a_2, h_2}$ in the network such that $h_1$ and $h_2$ are incompatible.

\item The $Abort\left( s \right)$ transition, for a switcher $s$, is enabled when the switcher has detected a deadlock. Its effect is to 
   set the variable $aborts\left[ s \right]$ to the maximum of the histories received from $a$, where $a$ is any acceptor. 
   The history $aborts'\left[ s \right]$ is then called an abort history.

 \item The $Init\left( s \right)$ transition, for a switcher $s$, always enabled, and its effect is to to pick a history $h$, set $inits\left[ s \right]$ to $h$, and to broadcast the message $\aseq{\quo{init}, h}$ to the acceptors. In $FGC_2$, composed with $FGC_1$, this history is constrained to be an abort history of $FGC_1$.
   The history $inits\left[ s \right]$ is then called an init history.

@@

The two rounds $FGC_1$ and $FGC_2$ are composed by requiring that $aborts_1 = inits_2$ at all times.
This model the switchers passing the histories $aborts\left[ s \right]$ from $FGC_1$ to $FGC_2$, where they broadcast to the acceptors and used in the $WakeUp\left( a \right)$ transitions to initialize the state of the acceptors.
This process can be repeated as often as needed, with new rounds $FGC_3$, $FGC_4$, etc., to try to make progress.

The initial $FGC$ algorithm also enters a deadlock whenever one acceptor crashes because the learners wait for the lower bound over all accepted histories of all acceptors to increase. However, the round-change mechanism is very resilient to faults: a single live acceptor with a single live switcher can abort the execution and change round.
To allow a new round to make progress after some acceptors crashed in a previous round, we introduce a component called the switching policy. With the Fast GC rounds, the task of the switching policy is to update the set of acceptors from one round to the next, ensuring that failed acceptors are removed or replaced by live ones. The switching policy must be queried by the switchers upon changing round to determine the new set of acceptors. This information can then be piggybacked to the
acceptors and learners, or the acceptors and learns can directly query the switching policy too. 
The set of switchers can be changed from round to round in the same fashion. Moreover, in an algorithm composed of several different types of rounds, the switching policy will also be responsible for choosing the type of round to execute next.

\section{Classic Generalized Consensus}

The Fast GC algorithm can process a proposal in two message delays when there is no contention on non-commuting commands. However, if there is
repeated contention on non-commuting commands, the algorithm may not make progress despite changing round, as the acceptors of the each new round
may be initialized with incompatible histories. The Fast Paxos algorithm \cite{Lamport06FastPaxos} is able to make progress under high contention by
designating a leader server that will totally order the proposals. However, to make our Fast GC algorithm work like Fast Paxos, we would have to modify it substantially. If we had spent effort testing Fast GC implementations and proving it correct, this effort would be wasted and we would have to redo all the tests and proofs from scratch.

Instead, we will now show how to obtain an algorithm with performance characteristics close to Fast Paxos, employing a leader to deal with contention, without modifying at all our existing Fast GC algorithm.

@@

Ring Paxos optimizes Chain Replication by having the leader broadcast
commands to the acceptors using IP multicast and only sending hashes
of the commands along the chain. Ring Paxos round can be modified to implement
conservative CGC instances in a similar way as for Chain Replication.

>>>>

\subsection{Fast Generalized Consensus (Aggrement By All?)}

In our first algorithm, Fast GC, learners each maintain a variable $hist\left[ l \right]$ which contain the currently accepted history of learner $l$.
For every learner $l$, $accepted\left[ l \right]$ is initialized to the empty history.
A proposer broadcasts its proposal to all the learners which unconditionally accept it by setting appending it to their $accepted\left[ l \right]$
variable and broadcasting to all the acceptors the new value of their accepted history.
Each acceptor $l$ maintains a variable $learned\left[ l \right]$, initially the empty history, containing its last learned history. A learner $l$
learns a new history $h$ when $h \geq learned\left[ l \right]$ and $h$ is a lower bound of the set consisting of the maximal accepted history received
from each acceptor.

@@

\item A $Learn\left( l, h \right)$ transition, for a learner $l$ and a history $h$, enabled when $h \geq learned\left[ l \right]$ and $h$ is a lower
bound of the set $\aset{MaxReceivedFrom\left( a \right) : a \in Acceptors}$, where $MaxReceivedFrom\left( a \right)$ is the maximal accepted history received
from the acceptor $a$.

@@

The algorithm Fast GC cannot learn new histories if two acceptors $a_1$ and $a_2$ have two histories which are not compatible because, since one can only append to the histories $accepted\left[ l \right]$, all the possible future lower bounds of the set $\aset{MaxReceivedFrom\left( a \right) : a \in Acceptors}$ are bounded above by $GLB\left( \aset{accepted\left[ a_1 \right], accepted\left[ a_2 \right]} \right)$.
This situation occurs when there is contention on non-commuting commands: for example, when two different proposers send two commands $c_1$ and $c_2$ that do not commute, an acceptors $a_1$ accepts $c_1$ before $c_2$, and an acceptor $a_2 \neq a_1$ accepts $c_2$ before $c_1$.

@@

We write $FGC_1$ for the first round of the Fast GC algorithm.
The rounds $FCG_1$ and $FCG_2$ each have their own copies of the arrays $accepted$ and $learned$, noted $accepted_1$, $accepted_2$, $learend_1$, and $learned_2$ when the round is not clear from the context. 
We assume that network messages are tagged with a round number and that a message tagged with $i\in\aset{1,2}$ can only be received in $FCG_i$, creating two isolated logical networks.

@@

We modify the FCG algorithm by introducing new transitions and modifying the $Accept\left( a \right)$ transition as follows.
We add the new variables $status\left[ a \right]$ for every acceptor $a$, which are initialized to $\quo{idle}$. The purpose of $status\left[ a \right]$ is to prevent the acceptor $a$ from accepting commands until its status changes from $\quo{idle}$ to $\quo{ready}$ and to stop it when its status goes to $\quo{stopped}$.
We also add the new variables $inits\left[ s \right]$ and $aborts\left[ s \right]$, for every switcher $s$, initialized to the empty history in $FGC_1$ and to $\quo{none}$ in $FGC_2$. The purpose of the arrays $aborts$ and $inits$ is to transfer state between $FGC_1$ and $FGC_2$. The arrays $aborts$ and $inits$ form what we call the \emph{inter-round} interface.
The algorithms $FGC_1$ and $FGC_2$ each have their own copy of the arrays $status$, $inits$, and $aborts$, and we adopt the same convention as above consisting in adding a subscript denoted the round to a variable when the round is not clear from the context.

\begin{compactitem}
\item We modify the $Accept\left( a \right)$ transition by enabling it only if $status\left[ a \right] = \quo{ready}$ and by having the acceptor $a$ additionally broadcast its new accepted value to the switchers.
Switchers can therefore detect that the algorithm is deadlocked when there are two broadcast messages $\aseq{\quo{accepted}, a_1, h_1}$ and $\aseq{\quo{accepted}, a_2, h_2}$ in the network such that $h_1$ and $h_2$ are incompatible.

\item The $Stop\left( s \right)$ transition, for a switcher $s$, is enabled when the switcher has detected a deadlock or has waited for too long without receiving any message from the acceptors. Its effect is to broadcast a $\aseq{\quo{stop}, s}$ message to the acceptors.

\item The $Stop\left( a \right)$ transition, for an acceptor $a$, is enabled when the acceptor $a$ can receive a $\aseq{\quo{stop}, s}$ message from a switcher $s$.
  Its effect is to update $status\left[ a \right]$ to $\quo{stopped}$ and to send the message $\aseq{\quo{stopped},a,accepted\left[ a \right]}$ to the switcher $s$.

\item The $Abort\left( s \right)$ transition, for a switcher $s$, is enabled when $aborts\left[ s \right] = \quo{none}$ and the switcher has received
at least on $\aseq{\quo{stopped}, a, h}$ message. Its effect is to set the variable $aborts\left[ s \right]$ to the history $h$ received from $a$. The
history $aborts'\left[ s \right]$ is then called an abort history. Because the switcher has received a $\aseq{\quo{stopped},a,h}$ message from the acceptor $a$, no history greater than $h$ can ever be learned anymore. Therefore, an abort history can be seen as a safe starting point for a new round to resume execution.

\item The $Init\left( s \right)$ transition, for a switcher $s$, is enabled when $inits\left[ s \right]=\quo{none}$, and its effect is to to pick a history $h$, set $inits\left[ s \right]$ to $h$, and to broadcast the message $\aseq{\quo{init}, h}$ to the acceptors. In $FGC_2$, composed with $FGC_1$, this history is constrained to be an abort history of $FGC_1$. The history $inits\left[ s \right]$ is then called an init history.

@@

The two rounds $FGC_1$ and $FGC_2$ are composed as follows
\begin{itemize}
  \item Requiring that a transition of the composition be a transition of either specification but not both, except for the $Abort\left( s \right)$ and $Init\left( s \right)$ transitions, which are allowed to take place at the same time. 
  \item Requiring that $aborts_1 = inits_2$ at all times.
\end{itemize}
This models two independent rounds being linked only by the switchers passing the histories $aborts\left[ s \right]$ from $FGC_1$ to $FGC_2$: switchers gather abort histories in $FGC_1$, then they broadcast them as init histories to the acceptors in $FGC_2$. The acceptors then initialize their state to an init histories in the $WakeUp\left( a \right)$ transition.
This process can be repeated as often as needed, with new rounds $FGC_3$, $FGC_4$, etc., to try to make progress learning new histories.

The initial $FGC$ algorithm also enters a deadlock whenever one acceptor crashes because the learners wait for the lower bound over all accepted histories of all acceptors to increase. However, the round-change mechanism is very resilient to faults and can abort anyway thanks to the timeout triggering the $Stop\left( s \right)$ transition: a single live acceptor with a single live switcher can abort the execution and change round.
To allow a new round to make progress after some acceptors crashed in a previous round, we introduce a component called the switching policy. With the Fast GC rounds, the task of the switching policy is to update the set of acceptors from one round to the next, ensuring that failed acceptors are removed or replaced by live ones. The switching policy must be queried by the switchers upon changing round to determine the new set of acceptors. This information can then be piggybacked to the
acceptors and learners, or the acceptors and learns can directly query the switching policy too. 
The set of switchers can be changed from round to round in the same fashion. Moreover, in an algorithm composed of several different types of rounds, the switching policy will also be responsible for choosing the type of round to execute next.

\section{The Simple-Leader Algorithm}

The Fast GC algorithm can process a proposal in two message delays when there is no contention on non-commuting commands. However, if there is
repeated contention on non-commuting commands, the algorithm may not make progress despite changing round, as the acceptors of the each new round
may be initialized with incompatible histories. The Fast Paxos algorithm \cite{Lamport06FastPaxos} is able to learn new histories under high contention by
designating a leader server that will totally order the proposals. However, to make our Fast GC algorithm work like Fast Paxos, we would have to modify it substantially. If we had spent effort testing Fast GC implementations and proving it correct, this effort would be wasted and we would have to redo all the tests and proofs from scratch.

Instead, we will now show how to obtain an algorithm with performance characteristics close to those of Fast Paxos, employing a leader to deal with contention, without modifying at all our existing Fast GC algorithm.
To do so, we define a new kind of rounds, the Simple-Leader rounds, which uses the same inter-round interface as the Agreement-By-All algorithm. 
The inter-round interface will allow use to combine Simple-Leader rounds and Agreement-By-All rounds in any order, obtaining an algorithm which can adapt its strategy to its environment, selecting the Agreement-By-All algorithm under low contention and switching to the more expensive Simple-Leader algorithm when there is contention.

In the Simple-Leader algorithm, the proposers send their proposals to a distinguished acceptor called the leader. 
The leader totally orders the proposals it receives and forwards them to all the acceptors which accept them in the order defined by the leader.
As in the Agreement-By-All algorithm, acceptors broadcast their accepted histories to the learners, which learn a new history when it is a lower bound of 
the set of $\aset{MaxReceivedFrom\left( a \right) : a \in Acceptors}$.
Thanks to its leader, a Simple-Leader round continues to learn new histories even when proposals conflict because the leader ensures that the accepted histories of the acceptors remain consistent.

The Simple-Leader algorithm treats all proposals equally and does not take advantage of commutativity. When the leader does not crash, each proposal is processed in three message delays. In comparison, in the absence of failures, the Agreement-By-All algorithm processes non-conflicting proposals in two message delays but can take arbitrarily long to process conflicting proposals.

To initialize a Simple-Leader round, the switchers send their init histories to the leader, which choose a unique one that it forwards to the acceptors, guaranteeing that all the acceptors are initialized with the same history even when conflicting init histories are sent concurrently.

The Simple-Leader algorithm can not learn new histories if its leader, or any other acceptor, crashes.
To recover from this deadlock we pass the baton to a new round, as in the Agreement-By-All algorithm, with a similar same mechanism: a switcher that times-out waiting for new messages from the acceptors broadcast a stop message, gathers the accepted history from at least one of the acceptors, and aborts.

Let us specify more precisely the Simple-Leader algorithm.

The specification of Simple-Leader algorithm uses the sets   
 $inits$, $aborts$, the arrays of variables $learned\left[ l \right]$,
$status\left[ a \right]$, and $accepted\left[ a \right]$, and two
additional arrays $accepted\left[ l \right]\left[ a \right]$ and
$stop\left[ s \right]\left[ a \right]$, where $l$ is a learner, $a$ is
an acceptor, and $s$ is a switcher.

Initially, the variables $inits$, $aborts$, and, for every
learner $l$, $learned\left[ l \right]$ are empty sets. For every
acceptor $a$, $status\left[ a \right]$ is ``idle'' and $hist\left[ a
\right]$ is the special value $None$. For every learner $l$ and acceptor $a$,
$accepted\left[ l \right]\left[ a \right]$ is the special value
$None$. For every switcher $s$ and acceptor $a$, $stopped\left[ s
\right]\left[ a \right]$ is set to $false$.  There are initially no
messages in the network.

Thanks to the inter-round interface, Simple-Leader rounds and Agreement-By-All rounds can be composed in any order.
To compose a Simple-Leader round $SL_1$ with an Agreement-By-All round $ABA_2$, we compose their specifications as before, adding the constraint that the set $aborts_1$ be equal to the set $inits_2$ at all times.

Informally, the composition is correct because abort histories are always safe in the sense that any of the abort histories could be the next learned history.
However not all can be learned and the initialization mechanism of the rounds ensures that at most one is chosen.
In the next section we present Composable Generalized Consensus, a specification of the behavior of rounds at the level of proposals, learned histories, init histories, and abort histories. Our composition theorem ensures that any set of rounds individually satisfying Composable Generalized Consensus can be composed in any order to yield an algorithm satisfying Composable Generalized Consensus.

@@

Ring Paxos optimizes Chain Replication by having the leader broadcast
commands to the acceptors using IP multicast and only sending hashes
of the commands along the chain. Ring Paxos rounds can be modified to implement
conservative CGC instances in a similar way as for Chain Replication.
